Current and future constraints on primordial magnetic fields

### 2017-02-06 pmf.tex
ready to go!

To calculate the Fisher matrix for a CMB experiment, we need to calculate partial derivatives of the CMB spectra and the bandpower covariance matrix. 
The partial derivatives are calculated numerically from CAMB spectra  \citep{lewis00}. 
We calculate the bandpower covariance matrix using the Knox formula \citep{knox97} for the survey parameters in Table~\ref{tab:experiments}. 
The sample variance contribution is calculated assuming the \planck{} best-fit \lcdm{} cosmology with a small gravitational wave contribution  ($r=0.01$) added. 
This is the black line in Fig.~\ref{fig:pmf-experiments}. 
We use a bandpower bin size of $\Delta\ell=50$. 
We  assume a 5\% uncertainty on the beam FWHM and a 1\% power calibration uncertainty.
In \S\ref{subsec:beamcal}, we test relaxing or tightening the beam and calibration uncertainty and find the beam and calibration uncertainties have a negligible impact on the PMF constraints.

@@

In this work, we restrict ourselves to power spectrum (i.e.~2-point estimators) searches for PMFs. 
Currently the power spectrum limits from \planck{} or \pb{} are better than the 4-point upper limits. 
Although outside the scope of this work, it would be interesting to extend this analysis to 4-point estimators in the future. 
First, the 4-point limits should improve faster as the noise level falls. 
Second, in the case of a detection, the 4-point estimators would almost certainly come into play to learn more about the vector modes of the PMFs. 
The detection (or non-detection) of the vector PMF signal in the 4-point estimators could also be used to argue for whether it is more likely that any observed tensor power is due to  inflationary gravitational waves or  PMFs. 
Finally, a lensing estimator might be used to `de-lense' the B-mode power spectrum, thereby suppressing the lensed B-mode signal and allowing better limits on PMFs. 
Such de-lensing techniques have long been proposed for inflationary gravitational wave searches \citep[e.g.][]{knox02,kesden02,seljak04a,simard15} and more recently been demonstrated on real \planck{} data \citep{larsen16}.

@@

The experiments that will begin taking data in 2017 will dramatically improve our knowledge of PMFs. 
In the minimal cosmological model, the 1-$\sigma$ forecast for \planck{}+\ho{} is $\sigma(\apmf)=\fisherAplk$ for a best-motivated, nearly-scale invariant PMF spectrum (i.e.~$n_B=-2.9$). 
This is on-par with, although 30\% better than, the actual upper limit from the \planck{} data.  
Adding  EE and BB bandpowers from two stage-III experiments, \sptnew{} and \simons{}, reduces the uncertainty by an order of magnitude in the 7-parameter model to $\sigma(\apmf)=0.015$. 
The relative improvement is larger in the more realistic 11-parameter model, as parameter degeneracies substantially weaken (by a factor of five) the \planck+\ho{} constraints on \apmf{}, while weakening the stage III constraint by a  more modest 40\%. 
Thus within the 11-parameter model however,  the addition of stage III CMB experiments improves the \apmf{} uncertainty by a factor of $\sim$\,35 to  $\sigma(\apmf)=0.022$. 
The parameter degeneracies largely disappear for steeper PMF spectra (i.e. $n_B=-1$ or 2) as the PMF B-mode spectra then peaks at very small scales and this small-scale power can not be mimicked by any of the other 11-parameters. 
The improvement from adding the SPT-3G and Simons Array experiments to \planck{} remains impressive for these values of $n_B$.
For the 11-parameter model and $n_B=-1$, adding SPT-3G and Simons Array to \planck{} leads to a 17-fold reduction in $\sigma(\apmf)$ from \fisherAlooseNbOneplk{} to \fisherAlooseNbOneSThree{}.
For $n_B=2$, there is a 20-fold reduction from $\sigma(\apmf) = \fisherAlooseNbTwoplk$ to  \fisherAlooseNbTwoSThree. 
We can expect substantially tighter constraints on PMFs by the end of the decade.

@@

The primary motivations behind the proposed CMB-S4 experiments are to search for inflationary gravitational waves and to measure the neutrino masses.
However, CMB-S4 would also enable extremely sensitive searches for PMFs. 
We begin by considering our fiducial CMB-S4 configuration, as laid out in Table~\ref{tab:experiments}. 
For this fiducial configuration and the 11-parameter model, we find a 3-fold reduction over the stage III experiments for all three PMF template shapes. 
The expected uncertainties for CMB-S4 are $\sigma(\apmf^{n_B=-2.9}) =   \fisherAlooseSFour$, $\sigma(\apmf^{n_B=-1}) =   \fisherAlooseNbOneSFour$, and $\sigma(\apmf^{n_B=2}) =   \fisherAlooseNbTwoSFour$. 
These represent more than a 50-fold improvement on current limits.

Given that CMB-S4 is being designed, it is worth considering how design decisions would affect the final PMF constraints. 
We look at five aspects of the experiment: the instrumental sensitivity (as reflected by the final noise level in the maps), the telescope size or beam FWHM,  the low-$\ell$ noise performance, the amount of sky surveyed,  and how well the beam shape and calibration must be known. 
Of these five aspects, we find the first two to be very important to PMF searches and the second two to be somewhat important. 
As might be expected, in some cases the optimal design depends on the shape of the PMF spectrum ($n_B$) as this sets  the relative power between large and small angular scales.  
However, as was discussed in \S\ref{sec:template},  there are theoretical reasons to expect the PMF spectrum to be close to a scale-invariant spectra. 
While for completeness we present results for three values of $n_B$, we recommend favoring the optimum for $n_B=-2.9$.

@@

\begin{figure}[htb]\centering
\includegraphics[width=0.45\textwidth,clip,trim={2.cm 12.5cm 11cm 7.5cm}]{pmf_beam.pdf}
  \caption[]{ \label{fig:beam}
 Larger telescopes (and thus smaller beams) improve limits on the PMF power. 
 Forecasts are shown as a function of the beam FWHM in arcminutes for each of the three PMF scalar indices: $n_B = -2.9$ (solid), -1 (dashed), and +2 (dash-dot). 
   In the $n_B=2$ case, the uncertainties have been multiplied by a factor of $10^5$ for plotting purposes. 
  The black lines are for the fiducial 11-parameter model, while the light blue lines are for the minimal \lcdm{}+\apmf{} model. 
  The cosmological model only matters in the nearly-scale invariant PMF case ($n_B=-2.9$), where the information on large angular scales dominates and the telescope size is relatively unimportant. 
  The plateau near 4$^\prime$ for $n_B=-1, +2$ is where the gain from the $\ell\sim3000$ peak is saturated, but the resolution isn't yet adequate to pick up the next peak at $\ell > 8000$ (see  Fig.~\ref{fig:pmf-nb}).  
    }
\end{figure}

@@

However, larger telescopes are also more expensive to build which means for a fixed experimental budget, they would necessitate less ambitious focal planes and lower instantaneous mapping speeds. 
We very crudely approximated a cost-neutral setup with a noise level of 1.3 \ukarcmin{} for a FWHM of 4$^\prime$; a noise level of 2.9 \ukarcmin{} for a FWHM of 2$^\prime$; and a noise level of 4.1 \ukarcmin{} for a FWHM of 1$^\prime$. 
We find that 
improving the mapping speed rather than telescope size leads to somewhat better potential limits for these experimental configurations, by a factor of 1.6, 1.2 or 1.1 for $n_B = -2.9, -1, +2$ respectively. 
However the results are close enough that the crudeness of the cost-neutral estimates used here could be a factor, and the preference might flip for more accurately costed setups. 
In short,  the tradeoff between larger telescopes or more detectors is largely a wash, although one might lean towards adding more detectors. 
%Raw sensitivity is important to an experiment's ability to study PMFs.

@@

A third question is how much sky to observe with CMB-S4. 
Unsurprisingly, we find opposing preferences for the nearly scale-invariant $n_B=-2.9$ (which peaks at large angular scales) and the bluer $n_B=-1$ or 2 spectra which peak at small scales. 
As shown in  Fig.~\ref{fig:area}, for $n_B=-2.9$, observing at least 15\% of the sky is very important to a search for nearly scale-invariant PMFs, and it is best  to cover as much sky as possible. 
A caveat to this analysis is that it is likely easier to remove galactic foregrounds to a specified level on targeted, `clean' patches as opposed to a substantial fraction of the sky, and we would expect galactic foregrounds to be important for the large angular scales. 
With that said, the PMF constraints improve by a factor of 1.4-1.5 (depending on the cosmological model) going from 10\% to 25\% of the full sky, and are nearly flat (a 6\% improvement) from 25\% to 70\% of the sky (the widest area likely to be possible after galactic cuts).
Conversely, $n_B=-1$ or +2, the optimal survey area is smaller at 10\% of the full sky. 
Note however that the minimum is extremely broad with only a 10\% worsening of the expected uncertainty as the survey area is increased from 10\% to 70\% of the  full sky. 
A survey of around 25\% of the sky would perform well for all three considered PMF models.

@@

We generally find the forecasts to be insensitive to the 1/f knee or galactic foreground removal at the default beam size of 4$^\prime$. 
It appears that even in the nearly scale-invariant case, the PMF constraint is being driven by smaller angular scales due to the the degeneracy with the tensor-to-scalar ratio at large angular scales. 
Without this degeneracy, i.e. in the simple \lcdm{}+\apmf{} model, dropping \lknee{} from 400 to 50 reduces the uncertainty by a factor of 3.2 for a nearly scale-invariant PMF. 
In this case, the degradation accelerates once the 1/f knee goes above $\ell \simeq 70$. 
This is consistent with the earlier statement for the same simple model that the current upper limits are dominated by the BICEP2/Keck Array data at low-$\ell$. 
For the bluer PMF spectra ($n_B=-1$ or 2), the forecasts show no dependence on the knee frequency. 
For the fiducial 11-parameter model, our ability to separate a PMF from other physics is coming from sub-degree angular scales. 
%An instrument built to precisely measure gravitational lensing of the CMB should do very well at searching for PMFs. 
The low frequency noise performance of CMB-S4 is not crucial to PMF searches.

@@

Finally, we consider if searches for PMFs introduce new requirements on the accuracy to which the beam or  calibration of an experiment must be known. 
We find they do not. 
We parameterize the beam uncertainty as a fractional uncertainty on the FWHM of the Gaussian beam, and calibration uncertainty as an overall power uncertainty. 
We find negligible, sub-percent shifts in the forecasted uncertainty for calibration uncertainties from 0.2 to 5\% and beam FWHM uncertainties from 2 to 12.5\%. 
The PMF constraints  for all three power law indices considered are insensitive to the beam and calibration uncertainties.

@@

In this work, we have improved the current upper limits on the strength of   primordial magnetic fields by including more CMB B-mode polarization data. 
By adding \bicepkeck{}, \pb, and \sptpol{} to \planck{} we find the 95\% CL upper limit on the PMF power falls nearly two-fold from $\apmf < \upperAplk$ to $\apmf < \upperAall$. 
The biggest contributor to this improvement is the low-$\ell$ data from \bicepkeck{}.

We have also shown that the next generation of experiments should dramatically reduce these limits, with the potential to detect PMFs for the first time. 
These so-called stage III experiments, which will begin taking data in 2017, can be expected to set upper limits at the level of $\apmf < 0.02$ even after marginalizing over a five-parameter extension to \lcdm{} and foregrounds.

The potential for detection increases even further with planned experiments like the Simons Observatory or CMB-S4. 
We show that an ideal version of CMB-S4 might decrease the 95\% CL upper limits eight-fold, to $\apmf < 0.0025$, and less optimistic versions can still set upper limits on order of $\apmf < 0.006$ in a 11-parameter cosmological model with  \lcdm{} +  $r$ + \nrun{} +  \neff{} + \mnu{} + \apmf. 
This represents a three-fold improvement over the forecasts for the stage III experiments and a more than 100-fold improvement over the current limits. 
We consider how the design of these experiments will impact the PMF limits, looking at the experimental sensitivity, angular resolution, survey area, low-frequency noise performance, and beam or calibration uncertainties. 
We find experiment's sensitivity followed by telescope size to be the most important factors in predicting the PMF limits. 
CMB-S4 will be a very exciting probe of PMFs and other sources of cosmic birefringence.

>>>>

To calculate the Fisher matrix for a CMB experiment, we need the bandpower covariance matrix and partial derivatives of the CMB spectra. 
The partial derivatives are calculated numerically from CAMB spectra  \citep{lewis00}. 
We calculate the bandpower covariance matrix using the Knox formula \citep{knox97} for the survey parameters in Table~\ref{tab:experiments}. 
The sample variance contribution is calculated assuming the \planck{} best-fit \lcdm{} cosmology with a small gravitational wave contribution  ($r=0.01$) added. 
This is the black line in Fig.~\ref{fig:pmf-experiments}. 
We use a bandpower bin size of $\Delta\ell=50$. 
We  assume a 5\% uncertainty on the beam FWHM and a 1\% power calibration uncertainty.
In \S\ref{subsec:beamcal}, we test relaxing or tightening the beam and calibration uncertainty and find the beam and calibration uncertainties have a negligible impact on the PMF constraints.

@@

In this work, we restrict ourselves to power spectrum (i.e.~2-point estimators) searches for PMFs. 
Currently the power spectrum limits from \planck{} or \pb{} are better than the 4-point upper limits. 
Although outside the scope of this work, it would be interesting to extend this analysis to 4-point estimators in the future. 
First, the 4-point limits should improve faster as the noise level falls. 
Second, in the case of a detection, the 4-point estimators would almost certainly come into play to learn more about the vector modes of the PMFs. 
The detection (or non-detection) of the vector PMF signal in the 4-point estimators could then be used to argue for whether it is more likely that any observed tensor power is due to  inflationary gravitational waves or  PMFs. 
Finally, a lensing estimator might be used to `de-lense' the B-mode power spectrum, thereby suppressing the lensed B-mode signal and allowing better limits on PMFs. 
Such de-lensing techniques have long been proposed for inflationary gravitational wave searches \citep[e.g.][]{knox02,kesden02,seljak04a,simard15} and more recently been demonstrated on real \planck{} data \citep{larsen16}.

@@

The experiments that will begin taking data in 2017 will dramatically improve our knowledge of PMFs. 
In the minimal cosmological model, the 1-$\sigma$ forecast for \planck{}+\ho{} is $\sigma(\apmf)=\fisherAplk$ for a best-motivated, nearly-scale invariant PMF spectrum (i.e.~$n_B=-2.9$). 
This is on-par with, although 30\% better than, the actual upper limit from the \planck{} data.  
Adding  EE and BB bandpowers from two stage-III experiments, \sptnew{} and \simons{}, reduces the uncertainty by an order of magnitude in the 7-parameter model to $\sigma(\apmf)=0.015$. 
The relative improvement is larger in the more realistic 11-parameter model, as parameter degeneracies substantially weaken (by a factor of five) the \planck+\ho{} constraints on \apmf{}, while weakening the stage III constraint by a  more modest 40\%. 
Thus within the 11-parameter model,  the addition of stage III CMB experiments improves the \apmf{} uncertainty by a factor of $\sim$\,35 to  $\sigma(\apmf)=0.022$. 
The parameter degeneracies largely disappear for steeper PMF spectra (i.e. $n_B=-1$ or 2) as the PMF B-mode spectra then peaks at very small scales and this small-scale power can not be mimicked by any of the other 11-parameters. 
The improvement from adding the SPT-3G and Simons Array experiments to \planck{} remains impressive for these values of $n_B$.
For the 11-parameter model and $n_B=-1$, adding SPT-3G and Simons Array to \planck{} leads to a 17-fold reduction in $\sigma(\apmf)$ from \fisherAlooseNbOneplk{} to \fisherAlooseNbOneSThree{}.
For $n_B=2$, there is a 20-fold reduction from $\sigma(\apmf) = \fisherAlooseNbTwoplk$ to  \fisherAlooseNbTwoSThree. 
We can expect substantially tighter constraints on PMFs by the end of the decade.

@@

The primary motivations behind the proposed CMB-S4 experiment are to search for inflationary gravitational waves and to measure the neutrino masses.
However, CMB-S4 would also enable extremely sensitive searches for PMFs. 
We begin by considering our fiducial CMB-S4 configuration, as laid out in Table~\ref{tab:experiments}. 
For this fiducial configuration and the 11-parameter cosmological model, we find a 3-fold reduction over the stage III experiments for all three PMF template shapes. 
The expected uncertainties for CMB-S4 are $\sigma(\apmf^{n_B=-2.9}) =   \fisherAlooseSFour$, $\sigma(\apmf^{n_B=-1}) =   \fisherAlooseNbOneSFour$, and $\sigma(\apmf^{n_B=2}) =   \fisherAlooseNbTwoSFour$. 
These represent more than a 50-fold improvement on current limits.

Given that CMB-S4 is being designed, it is worth considering how design decisions would affect the final PMF constraints. 
We look at five aspects of the experiment: the instrumental sensitivity (as reflected by the final noise level in the maps), the telescope size or beam FWHM,  the low-$\ell$ noise performance, the amount of sky surveyed,  and how well the beam shape and calibration must be known. 
Of these five aspects, we find the first two to be very important to PMF searches and the second two to be somewhat important. 
As might be expected, in some cases the optimal design depends on the shape of the PMF spectrum ($n_B$) as this sets  the relative power between large and small angular scales.  
However, as was discussed in \S\ref{sec:template},  there are theoretical reasons to expect the PMF spectrum to be close to a scale-invariant spectra. 
While for completeness we present results for three values of $n_B$, we recommend favoring the optima for $n_B=-2.9$.

@@

\begin{figure}[htb]\centering
\includegraphics[width=0.45\textwidth,clip,trim={2.cm 12.5cm 11cm 7.5cm}]{pmf_beam.pdf}
  \caption[]{ \label{fig:beam}
 Larger telescopes (and thus smaller beams) improve limits on the PMF power. 
 Forecasts are shown as a function of the beam FWHM in arcminutes for each of the three PMF scalar indices: $n_B = -2.9$ (solid), -1 (dashed), and +2 (dash-dot). 
   In the $n_B=2$ case, the uncertainties have been multiplied by a factor of $10^5$ for plotting purposes. 
  The black lines are for the fiducial 11-parameter model, while the light blue lines are for the minimal \lcdm{}+\apmf{} model. 
  The cosmological model only matters in the nearly-scale invariant PMF case ($n_B=-2.9$), where the information on large angular scales dominates and the telescope size becomes less important. 
  The plateau near 4$^\prime$ for $n_B=-1, +2$ is where the gain from the $\ell\sim3000$ peak is saturated, but the resolution isn't yet adequate to pick up the next peak at $\ell > 8000$ (see  Fig.~\ref{fig:pmf-nb}).  
    }
\end{figure}

@@

However, larger telescopes are also more expensive to build which means for a fixed experimental budget, they would necessitate less ambitious focal planes and lower instantaneous mapping speeds. 
We very crudely approximated a cost-neutral setup with a noise level of 1.3 \ukarcmin{} for a FWHM of 4$^\prime$; a noise level of 2.9 \ukarcmin{} for a FWHM of 2$^\prime$; and a noise level of 4.1 \ukarcmin{} for a FWHM of 1$^\prime$. 
We find that 
improving the mapping speed rather than telescope size leads to somewhat better potential limits for these experimental configurations, by a factor of 1.6, 1.2 or 1.1 for $n_B = -2.9, -1, +2$ respectively. 
However the results are close enough that the crudeness of the cost-neutral estimates used here is a worry, and the preference might flip for more accurately costed setups. 
In short,  the tradeoff between larger telescopes or more detectors is largely a wash, although one might lean towards adding more detectors. 
%Raw sensitivity is important to an experiment's ability to study PMFs.

@@

A third question is how much sky to observe with CMB-S4. 
Unsurprisingly, we find opposing preferences for the nearly scale-invariant $n_B=-2.9$ (which peaks at large angular scales) and the bluer $n_B=-1$ or 2 spectra which peak at smaller scales. 
As shown in  Fig.~\ref{fig:area}, observing at least 15\% of the sky is very important to a search for nearly scale-invariant PMFs, and it is best  to cover as much sky as possible. 
A caveat to this analysis is that it is likely easier to remove galactic foregrounds to a specified level on targeted, `clean' patches as opposed to a substantial fraction of the sky, and we would expect galactic foregrounds to be important for the large angular scales. 
With that said, the PMF constraints improve by a factor of 1.4-1.5 (depending on the cosmological model) going from 10\% to 25\% of the full sky, and are nearly flat (a 6\% improvement) from 25\% to 70\% of the sky (the widest area likely to be possible after galactic cuts).
Conversely for $n_B=-1$ or +2, the optimal survey area is smaller at 10\% of the full sky. 
Note however that the minimum is extremely broad with only a 10\% worsening of the expected uncertainty as the survey area is increased from 10\% to 70\% of the  full sky. 
A survey of around 25\% of the sky would perform well for all three considered PMF models.

@@

We generally find the forecasts to be insensitive to the 1/f knee or galactic foreground removal at the default beam size of 4$^\prime$. 
It appears that even in the nearly scale-invariant case, the PMF constraint is coming primarily from smaller angular scales due to the the degeneracy with the tensor-to-scalar ratio at large angular scales. 
Without this degeneracy, i.e. in the simple \lcdm{}+\apmf{} model, dropping \lknee{} from 400 to 50 reduces the uncertainty by a factor of 3.2 for a nearly scale-invariant PMF. 
In this case, the degradation accelerates once the 1/f knee rises to $\lknee{} \gtrsim 70$. 
This is consistent with the earlier statement for the same simple model that the current upper limits are dominated by the BICEP2/Keck Array data at low-$\ell$. 
For the bluer PMF spectra ($n_B=-1$ or 2), the forecasts show no dependence on the knee frequency. 
For the fiducial 11-parameter model, our ability to separate a PMF from other physics is coming from sub-degree angular scales. 
%An instrument built to precisely measure gravitational lensing of the CMB should do very well at searching for PMFs. 
The low frequency noise performance of CMB-S4 is not crucial to PMF searches.

@@

Finally, we consider if searches for PMFs introduce new requirements on the accuracy to which the beam or  calibration of an experiment must be known. 
We find they do not. 
We parameterize the beam uncertainty as a fractional uncertainty on the FWHM of s Gaussian beam, and calibration uncertainty as an overall power uncertainty. 
We find negligible, sub-percent shifts in the forecasted PMF uncertainties for calibration uncertainties from 0.2 to 5\% and beam FWHM uncertainties from 2 to 12.5\%. 
The PMF constraints  for all three power law indices considered are insensitive to the beam and calibration uncertainties.

@@

In this work, we have improved the current upper limits on the strength of   primordial magnetic fields by including more CMB B-mode polarization data. 
By adding \bicepkeck{}, \pb, and \sptpol{} to \planck{} we have found the 95\% CL upper limit on the PMF power falls nearly two-fold from $\apmf < \upperAplk$ to $\apmf < \upperAall$. 
The biggest contributor to this improvement is the low-$\ell$ B-mode data from \bicepkeck{}.

We have also shown that the next generation of experiments should dramatically reduce these limits, with the potential to detect PMFs for the first time. 
The so-called stage III experiments, which will begin taking data in 2017, can be expected to set upper limits at the level of $\apmf < 0.02$ even after marginalizing over a five-parameter extension to \lcdm{} and foregrounds.

The potential for detection increases even further with planned experiments like the Simons Observatory or CMB-S4. 
We have shown that an ideal version of CMB-S4 might decrease the 95\% CL upper limits eight-fold, to $\apmf < 0.0025$, and less optimistic versions can still set upper limits on order of $\apmf < 0.006$ in a 11-parameter cosmological model with  \lcdm{} +  $r$ + \nrun{} +  \neff{} + \mnu{} + \apmf. 
This represents a three-fold improvement over the forecasts for the stage III experiments and a more than 100-fold improvement over the current limits for this cosmological model. 
We have considered how the design of future experiments will impact the resulting PMF limits, looking at the experimental sensitivity, angular resolution, survey area, low-frequency noise performance, and beam or calibration uncertainties. 
We have found an experiment's sensitivity followed by telescope size to be the most important factors in predicting the PMF limits. 
CMB-S4 will be a very exciting probe of PMFs and other sources of cosmic birefringence.
